---
title: "loading-message-datatable.Rmd"
subtitle: "Display loading & updating messages in DataTables, see https://thomasswilliams.github.io/development/r/2022/09/16/r-markdown-loading-message-datatable.html"
author: "Thomas Williams <https://github.com/thomasswilliams>"
output: html_document
runtime: shiny
---

The two DataTables below, populated with random values, take the same amount of time to load and reload.

The second DataTable displays a "loading" message when empty and an "updating" message when reloading. This gives useful feedback to users that something is happening.

Click the "reload" button below to see the updating message.

``` {r global, echo = FALSE, message = FALSE, warning = FALSE}
# DataTable
# https://rstudio.github.io/DT/
library(DT)
```

```{r, echo = FALSE}
# refresh button
# note icons need "icon()" as per https://shiny.rstudio.com/reference/shiny/0.14/icon.html
shiny::actionButton("refreshbutton", "Reload", icon = shiny::icon("arrows-rotate"))
```

<!-- basic two-column layout, start outer div and define multiple columns, adapted from
     https://bookdown.org/yihui/rmarkdown-cookbook/multi-column.html -->
<div style="display: grid; grid-template-columns: 1fr 1fr; grid-column-gap: 10px; margin-top: 2em;">
<!-- start first column -->
<div>
```{r, echo = FALSE}
# output placeholder for table 1
DT::dataTableOutput("table1", width = "100%", height = "auto")
```
<!-- end first column -->
</div>
<!-- start second column -->
<div>
```{r, echo = FALSE}
# output placeholder for table 2
DT::dataTableOutput("table2", width = "100%", height = "auto")
```
<!-- end second column -->
</div>
<!-- end outer div -->
</div>

```{r, echo = FALSE}
# random numbers 0-20 as per https://stackoverflow.com/a/19352289
# 2 columns, 20 records
df1 <- data.frame(replicate(2, sample(0:20, 20, rep = TRUE)))
df2 <- data.frame(replicate(2, sample(0:20, 20, rep = TRUE)))

# render the simplest possible datatable (same for both "table1" and "table2", just
# for demo purposes)
output$table1 <- DT::renderDataTable({
  # render function is dependent on "refresh" button, so when refresh button is
  # clicked, table will be re-rendered (this code will be run)
  input$refreshbutton

  # sleep for 3 seconds
  Sys.sleep(3)

  # return datatable from dataframe
  DT::datatable(
    df1
  )
})

output$table2 <- DT::renderDataTable({
  input$refreshbutton
  Sys.sleep(3)
  DT::datatable(
    df2
  )
})
```

The loading and updating messages use the DataTable name and the following CSS:

```{css}
/* style body color to better differentiate from datatables */
body {
  background-color: #fafafb;
}

/* datatables div */
#table1, #table2 {
  background-color: #ffffff;
}

/* display message in empty DT datatable
   this only occurs on initial load of datatable (subsequent loads leave the table
   in place, styled with "recalculating" class to make the table inactive)
   "table2" div elements will be empty, use CSS to display message */
#table2:empty::after {
  content: "Loading (please wait)...";
  opacity: 0.5;
  font-style: italic;
}

/* when table has the "recalculating" class applied, make the table less prominent */
#table2.recalculating {
  opacity: 0.1 !important;
  /* need to position relatively so we can center "Updating" text */
  position: relative;
}

/* display "Updating" message when table reloads (for example, on parameter change)
   will be used when the table is already displayed */
#table2.recalculating::before {
  content: "Updating (please wait)...";
  font-style: italic;
  /* text should sit above table controls */
  position: absolute;
  /* centered above table as per https://stackoverflow.com/a/50958847 */
  top: 50%;
  left: 50%;
  transform: translate(-50%,-50%);
  /* make text larger */
  font-size: 1.6em;
  /* content above controls */
  z-index: 1000;
}
```